Reconstructing the Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-11-12 doc/paper-lcr.tex
Edited introduction

Along lines of sight containing strong gravitational lenses, the
perturbative effects of line of sight mass structure have been found to
be particularly important, with foreground and background structures
having a significant effect on the inferred lensing cross-section
\citep[\eg][]{WongEtal2012} and distance ratios
\citep[][]{DalalEtal2005}. Indeed, \citet{SuyuEtal2010} found that in
time delay lens cosmography the so-called ``external convergence''
$\kappa$ due to mass structures along the (unusually over-dense) line
of sight to the quadruply-imaged radio source B1608$+$656 was the
dominant source of uncertainty in their 5\% measurement of the time delay distance. The external shear can potentially be recovered in the
mass-modelling of the lens, but the convergence is undetectable from the
image positions, shapes and relative fluxes; this is the famous {\emph{ 
mass sheet-degeneracy}} \citep[see e.g.][for details]{FalcoEtal1985}.

@@

In order to propogate the effect of line-of-sight perturbations into 
uncertainties on time-delay distances we require the probability density function (PDF) for the external convergence
given our available data $\data$, $\Pr(\xkappa|\data)$
% PHIL's REVIEWS OF THE LITRATURE
\comments{\citet{GunnarssonEtal2006} investigated a related PDF,  for the
magnification, using a galaxy halo model. Applying empirical galaxy
scaling relations both to simulate mock catalogs and then to reconstruct
the line of sight mass distribution. While they did not explore highly
overdense lines of sight, or the impact of groups and clusters, they
found that under their simplifying assumptions, the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gamma|\data)$ from their survey data, running similar Monte Carlo
reconstructions of the mass in nine lens fields based on
\citeauthor{WilliamsEtal2006} and \citeauthor{MomchevaEtal2006}'s
photometric and spectroscopic measurements of galaxies close to the line
of sight. They then  compared the resulting predicted shear
distributions with the external shear demanded by the strong lens model.
Their reconstructions showed most of the shear to have been generated by
bright galaxies within 2~arcminutes of the lens, and that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear. They found significant
discrepancies between the lens model shear and the Monte Carlo
predictions, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies.
%
\citet{KarpenkaEtal2012} used a halo model to predict deterministically 
the magnification due to
galaxies along the line of sight to type Ia supernovae, and inferred the
parameters of scaling relations between light and mass in these galaxies
from hundreds of supernova lightcone simultaneously. Following
\citet{JonssonEtal2010}, they subtract off the mean convergence to ensure that
the universal mass budget is balanced, and compare models where the halos are
all have either truncated singular isothermal profiles, or 
NFW profiles. Both groups were able to 
detect lensing effects in the supernova data this way.
}
% TOM's DISTILLING OF SAID REVIEWS.
Previous investigation of line-of-sight perturbations have focused on 
magnification and shear.\citet{GunnarssonEtal2006} used a galaxy halo model combined with simple scaling relations to reconstruct
the line of sight mass distribution and found that the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two. Conversely \citet{KarpenkaEtal2012} used the brightness dispersion in type Ia supernovae to infer the parameters of both their galaxy halo model and mass-to-light scaling relation. Both \citeauthor{KarpenkaEtal2012} and \citet{JonssonEtal2010} were able to 
detect lensing effects in the supernova data this way.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gammax|\data)$ from their survey data,
\citep{WilliamsEtal2006;MomchevaEtal2006} of galaxies close to the line
of sight. Comparing the predicted shear
distributions with the external shear demanded by the strong lens model 
they found significant
discrepancies, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies. \citeauthor{WongEtal2011} also found that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear.

@@

Strong lenses, type Ia supernovae and other cosmologically interesting objects
are never viewed unlensed: 
there is some mass along the line of sight to them coming from intervening
structures. These structures may be physically associated with the target
e.g. other members of a group or cluster, or they may be the result of
chance alignments at any redshift.  
% While it is rare for three galaxies
% to line up well enough for both of the background sources to be strongly
% lensed \citep{GavazziEtal2008,CollettEtal2012a}, 
The large size of dark
matter halos makes partial alignments -- such that they
act as perturbing weak lenses --  a near certainty
\citep{Vale+White2003,HilbertEtal2007}. These additional structures
induce an external shear which can be recovered in the mass-modelling of
a strong lens, but they also introduce a convergence that is undetectable even
in these systems:
the lensed image positions, shapes and relative fluxes are affected, but not
in an unambiguous way
\citep{FalcoEtal1985}.

>>>>

Along lines of sight containing strong gravitational lenses, the
perturbative effects of line of sight mass structure have been found to
be particularly important, with foreground and background structures
having a significant effect on the inferred lensing cross-section
\citep[\eg][]{WongEtal2012} and distance ratios
\citep[][]{DalalEtal2005}. Indeed, \citet{SuyuEtal2010} found that in
time delay lens cosmography the so-called ``external convergence''
$\kappax$ due to mass structures along the (unusually over-dense) line
of sight to the quadruply-imaged radio source B1608$+$656 was the
dominant source of uncertainty in their 5\% measurement of the time delay distance. 
Whilst it is rare for three galaxies to line up well enough for both of
the background sources to be strongly lensed \citep{GavazziEtal2008,CollettEtal2012a}, 
the large size of dark matter halos makes partial alignments -- such that they
act as perturbing weak lenses --  a near certainty. The large scale of the
perturbers means the external lensing perturbations can be approximated
by a quadrupole lens characterised only by external convergence and shear.
The external shear can potentially be recovered in the
mass-modelling of the lens, but the convergence is undetectable from the
image positions, shapes and relative fluxes; this is the famous {\emph{ 
mass sheet-degeneracy}} \citep[see e.g.][for details]{FalcoEtal1985}.

@@

In order to propogate the effect of line-of-sight perturbations into 
uncertainties on time-delay distances we require the probability density function (PDF) for the external convergence
given our available data $\data$, $\Pr(\xkappa|\data)$
% PHIL's REVIEWS OF THE LITRATURE
\comments{\citet{GunnarssonEtal2006} investigated a related PDF,  for the
magnification, using a galaxy halo model. Applying empirical galaxy
scaling relations both to simulate mock catalogs and then to reconstruct
the line of sight mass distribution. While they did not explore highly
overdense lines of sight, or the impact of groups and clusters, they
found that under their simplifying assumptions, the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gammax|\data)$ from their survey data, running similar Monte Carlo
reconstructions of the mass in nine lens fields based on
\citeauthor{WilliamsEtal2006} and \citeauthor{MomchevaEtal2006}'s
photometric and spectroscopic measurements of galaxies close to the line
of sight. They then  compared the resulting predicted shear
distributions with the external shear demanded by the strong lens model.
Their reconstructions showed most of the shear to have been generated by
bright galaxies within 2~arcminutes of the lens, and that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear. They found significant
discrepancies between the lens model shear and the Monte Carlo
predictions, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies.
%
\citet{KarpenkaEtal2012} used a halo model to predict deterministically 
the magnification due to
galaxies along the line of sight to type Ia supernovae, and inferred the
parameters of scaling relations between light and mass in these galaxies
from hundreds of supernova lightcone simultaneously. Following
\citet{JonssonEtal2010}, they subtract off the mean convergence to ensure that
the universal mass budget is balanced, and compare models where the halos are
all have either truncated singular isothermal profiles, or 
NFW profiles. Both groups were able to 
detect lensing effects in the supernova data this way.
}
% TOM's DISTILLING OF SAID REVIEWS.
Previous investigation of line-of-sight perturbations have focused on 
magnification and shear.\citet{GunnarssonEtal2006} used a galaxy halo model combined with simple scaling relations to reconstruct
the line of sight mass distribution and found that the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two. Conversely \citet{KarpenkaEtal2012} used the brightness dispersion in type Ia supernovae to infer the parameters of both their galaxy halo model and mass-to-light scaling relation. Both \citeauthor{KarpenkaEtal2012} and \citet{JonssonEtal2010} were able to 
detect lensing effects in the supernova data this way.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gammax|\data)$ from their survey data,
\citep{WilliamsEtal2006;MomchevaEtal2006} of galaxies close to the line
of sight. Comparing the predicted shear
distributions with the external shear demanded by the strong lens model 
they found significant
discrepancies, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies. \citeauthor{WongEtal2011} also found that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear.

@@

Strong lenses, type Ia supernovae and other cosmologically interesting objects
are never viewed unlensed: 
there is some mass along the line of sight to them coming from intervening
structures. These structures may be physically associated with the target
e.g. other members of a group or cluster, or they may be the result of
chance alignments at any redshift.

\citep{Vale+White2003,HilbertEtal2007}. These additional structures
induce an external shear which can be recovered in the mass-modelling of
a strong lens, but they also introduce a convergence that is undetectable even
in these systems:
the lensed image positions, shapes and relative fluxes are affected, but not
in an unambiguous way
\citep{FalcoEtal1985}.
